---
title: "Stats Rethinking"
author: "Coralie Williams"
execute:
  echo: false
format:
  html:
    toc: true
    toc-location: body 
    code-fold: true
    code-summary: "Show the code"
editor: visual
---

[\[Statistical Rethinking 2nd edition book pdf\]](https://github.com/Booleans/statistical-rethinking/blob/master/Statistical%20Rethinking%202nd%20Edition.pdf)

```{r include=F}
knitr::opts_chunk$set(message = FALSE)
```

DAG mermaid diagrams testing

```{r}
library(DiagrammeR)
mermaid("
graph LR
    A-->B
")
```

------------------------------------------------------------------------

## Chapter 14: Adventures in Covariance

### Example datasets

-   Cafe wait time

-   Prosocial chimpanzee experiment

-   Dyad household gifts ([Koster & Leckie, 2014](https://doi.org/10.1016/j.socnet.2014.02.002))

-   Oceanic tools

-   Primate phylogenetic analysis

------------------------------------------------------------------------

### 1. Varying slopes by construction

```{r}
library(brms)
library(dplyr)
library(tidyverse)


#### Simulate the population
a       <-  3.5  # average morning wait time
b       <- -1    # average difference afternoon wait time
sigma_a <-  1    # std dev in intercepts
sigma_b <-  0.5  # std dev in slopes
rho     <- -.7   # correlation between intercepts and slopes

# the next three lines of code simply combine the terms, above
mu     <- c(a, b)

cov_ab <- sigma_a * sigma_b * rho
sigma  <- matrix(c(sigma_a^2, cov_ab, 
                   cov_ab, sigma_b^2), ncol = 2)


sigmas <- c(sigma_a, sigma_b)          # standard deviations
rho    <- matrix(c(1, rho,             # correlation matrix
                   rho, 1), nrow = 2)

# now matrix multiply to get covariance matrix
sigma <- diag(sigmas) %*% rho %*% diag(sigmas)

# how many cafes would you like?
n_cafes <- 20

set.seed(5)  # used to replicate example

vary_effects <- 
  MASS::mvrnorm(n_cafes, mu, sigma) %>% 
  data.frame() %>% 
  set_names("a_cafe", "b_cafe")


#### Simulate observations 
n_visits <- 10
sigma    <-  0.5  # std dev within cafes

set.seed(22)  # used to replicate example

d <-
  vary_effects %>% 
  mutate(cafe = 1:n_cafes) %>% 
  expand_grid(visit = 1:n_visits) %>% 
  mutate(afternoon = rep(0:1, times = n() / 2)) %>% 
  mutate(mu = a_cafe + b_cafe * afternoon) %>% 
  mutate(wait = rnorm(n = n(), mean = mu, sd = sigma)) %>% 
  select(cafe, everything())


#### View dataset
d %>%
  mutate(afternoon = ifelse(afternoon == 0, "M", "A"),
         day       = rep(rep(1:5, each = 2), times = n_cafes)) %>%
  filter(cafe %in% c(3, 5)) %>%
  mutate(cafe = str_c("café #", cafe)) %>% 
  
  ggplot(aes(x = visit, y = wait, group = day)) +
  geom_point(aes(color = afternoon), size = 2) +
  geom_line(color = "#8B9DAF") +
  scale_color_manual(values = c("#80A0C7", "#EEDA9D")) +
  scale_x_continuous(NULL, breaks = 1:10, labels = rep(c("M", "A"), times = 5)) +
  scale_y_continuous("wait time in minutes", limits = c(0, NA)) +
  theme_pearl_earring(axis.ticks.x = element_blank(),
                      legend.position = "none") +
  facet_wrap(~ cafe, ncol = 1)
```

------------------------------------------------------------------------

### 2. Advanced varying slopes: Household dyads dataset

Prosocial chimpanzee experiment

```{r}
# Load chimpanzees prosocial experiement data
data(chimpanzees, package = "rethinking")
d <- chimpanzees
rm(chimpanzees)

# wrangle
d <-
  d %>% 
  mutate(actor     = factor(actor),
         block     = factor(block),
         treatment = factor(1 + prosoc_left + 2 * condition),
         # this will come in handy, later
         labels    = factor(treatment,
                            levels = 1:4,
                            labels = c("r/n", "l/n", "r/p", "l/p")))

glimpse(d)
```

------------------------------------------------------------------------

### 3. Instruments and causal design

/

------------------------------------------------------------------------

### 4. Social relations as correlated varying effects

Household dyads:

-   25 households

-   300 dyads (social networks)

-   600 gifts observations

```{r}
# Load librairies
library(rethinking)

# Load dataset of household dyads
data(KosterLeckie)
```

"The variables `hidA` and `hidB` tell us the household IDs in each dyad, and `did` is a unique dyad ID number" (p. 462). To get a sense of the interrelation among those three ID variables, we'll make a tile plot.

**Simple model**

```{r}
# Load librairies
library(rethinking)
library(brms)

# Fit simple brm model: discussed during stats book club
model.dyad.brm <- brm(giftsAB ~ 1 + (1|hidA) + (1|hidB) + (1|did), data=kl_dyads)

```

**Multi-predictor model**

Fit model in brms

```{r}


```

The `echo: false` option disables the printing of code (only output is displayed).

Notes:

-   brms is not set up to allow covariances among distinct random effects with the same levels: https://github.com/ASKurz/Statistical_Rethinking_with_brms_ggplot2_and_the_tidyverse_2\_ed/issues/6

### 5. Continuous categories and the Gaussian process

There is a way to apply the varying effects approach to continuous categories... The general approach is known as **Gaussian process regression**. This name is unfortunately wholly uninformative about what it is for and how it works.

> Gaussian process models constitute a class of probabilistic statistical models in which a Gaussian Process (GP) is used to described the Bayesian "a prior" uncertainty about a latent function.

#### 

Resources/Links for Gaussian Processes:

https://katbailey.github.io/post/gaussian-processes-for-dummies/
